---
title: "COVID rates vs. Election Results in NYC"
subtitle: "DATA 607 Final Project"
author: 
  - Shana Green
  - Mark Gonsalves 
  - Dominika Markowska-Desvallons
  - Orli Khaimova 
  - John Mazon
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tigris_use_cache = TRUE)
```

### Introduction

As a group, we worked with public CSV data from NYC Health Dept relating to 
positive cases by zip code in New York City. Secondly, we utilized our shared 
GitHub to upload a CSV with information regarding Presidential Voting Results by 
election district which we were able to find [here]("https://vote.nyc/page/election-results-summary"). 

We were able to communicate via phone call, text message, email, zoom meeting and
collaborating with shared group [Github]("https://github.com/data607project3-SMDOJ/finalproject").

In early October, New York announced a zoning strategy to tackle on the rising 
COVID-19 cases. However, once the unofficial results from the 2020 Presidential 
election came out, we noticed that there was a resemblance visually. There were 
similar patterns between areas that were marked in the red zone, meaning a 7 day 
positivity rate of over 3%, and areas where the majority voted for one candidate. 
We wanted to further see if that was just a visual resemblance or if there is a 
correlation.

We wanted a project that would be focused on current events and Covid and the
Presidential election are about as topical as you can get.  We thought combining
the Covid rate in New York City with the voting patterns from the 2020 Presidential
elections would be interesting to see if there is a correlation between voting for
a candidate and the current rate of Covid spread in New York City.  We used COVID
data for the week leading up to and including election day.


```{r, echo=FALSE, out.width = "50%", fig.align = "center"}
image = "https://i.redd.it/gkwnuqxxpor51.jpg"

```

<center><img src="`r image`"></center>

### Loading Libraries

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(magrittr)
library(sf)
library(tmaptools)
library(tmap)
library(raster)
library(zipcodeR)
library(tigris)
```

### Loading Data

The `covid` data is uploaded directly from the NYC Department of Health and is updated daily while the election data was downloaded from [vote.nyc]("https://vote.nyc/page/election-results-summary". We had to update the link for the `covid` dataset to keep the variable names constant and avoid errors later on.

```{r}
#covid <- read.csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/latest/pp-by-modzcta.csv")

covid <- read.csv("https://raw.githubusercontent.com/data607project3-SMDOJ/finalproject/main/Data/pp-by-modzcta.csv")

election <- read.csv("https://raw.githubusercontent.com/data607project3-SMDOJ/finalproject/main/Data/00000100000Citywide%20President%20Vice%20President%20Citywide%20EDLevel.csv",
                     header = FALSE)
```

### Cleaning Data

As with many other projects we utilized tidyverse functions to sort the data. 
we used `filter` to grab information with an end date of November 3, 2020(election day).
We made the data set longer by putting zipcodes in a column utilizing `pivot_longer` 
function. We also cleaned up our column names utilizing `set_colnames` to make it more 
understandable with terms such as `date`, `zipcode`, `rate`  

We also performed routine maintenance of removing leading “x” from the zip code
headers utilizing `str_remove` and `str_replace` to remove punctuation from numbers.

We also cleaned the `election` data. We removed the first 11 columns as they were
messy and actually the headers. `AD` is the assembly district and `ED` is the 
election district. We also filtered out the results to see where the EDAD was active
as some were combined with other election districts since they were very small. We 
also filtered out the candidates and only looked at the Democratic and Republican 
candidates, Joseph R. Biden and Donald J. Trump. Afterwards were replaced the entire
string, which included the presidential and vice-presidential candidate as a unit,
with just the last name of the presidential candidate. We also combined the tallies
for each candidate as Trump was a candidate for the Republican and Conservatives 
parties and Biden was a candidate for the Democratic and Working Families parties.
Afterwards, we found the proportion of votes for the candidates, excluding the 
independent party candidates, in each election district. 

We also had to create a `ADED` variable which is a 5 digit code which would be 
used later on to perform a`left_join`.


```{r, warning = FALSE, message = FALSE}
# filtered election day, made the dataset longer by putting zip codes in a column
covid <- covid %>%
  filter(End.date == "11/03/2020") %>%
  pivot_longer(2:184, names_to = "zipcode", values_to = "rate") %>%
  set_colnames(c("date", "zipcode", "rate"))

# cleaned up the zip code variable by removing extra characters
covid$zipcode <- covid$zipcode %>%
  str_remove("X") %>%
  str_replace("\\.", " ")

zipcodes <- covid[(7:183), ] %>%
  dplyr::select(zipcode) %>%
  as.vector()
 
# extracted columns, filtered out ED that were not combined, extracted more columns, 
# replaced the candidate names with their last name
# made tally numeric
# combined tallies for each candidate since diff party
election <- election %>%
  dplyr::select(12:22) %>%
  set_colnames(c("AD", "ED", "County", "EDAD_Status", "Event", 
               "Party/Independent_Body", "Office/Position Title", "District_Key",
               "VoteFor", "Candidate", "Tally")) %>%
  filter(EDAD_Status == "IN-PLAY") %>%
  dplyr::select(c("AD", "ED", "County", "Candidate", "Tally")) %>%
  filter(str_detect(Candidate, "Trump|Biden")) %>%
  mutate(Candidate = str_replace_all(Candidate, ".*Trump.*", "Trump"),
         Candidate = str_replace_all(Candidate, ".*Biden.*", "Biden"),
         Tally = as.numeric(str_replace_all(Tally, ",", ""))) %>%
  group_by(AD, ED, Candidate) %>%
  mutate(Tally = sum(Tally)) %>%
  ungroup() %>%
  unique() %>%
  group_by(AD, ED) %>%
  mutate(Proportion = Tally / sum(Tally)) %>%
  filter(sum(Tally) != 0) %>%
  ungroup() %>% 
  mutate(ED = as.character(ED),
         AD = as.character(AD),
         ADED = ifelse(str_length(ED) == 1, str_c(AD, "00", ED),
                if_else(str_length(ED) == 2, str_c(AD, "0", ED), 
                str_c(AD, ED))),
         ADED = as.numeric(ADED)) 



```

### Matching Zip Codes to Election District

Here we used shapefiles in order to match the election district to the zip code. We had some challenges here but eventually we figured it out. We downloaded the shapefiles from NYC OpenData for the modified 
ZIP Code Tabulation Areas (ZCTA) and election districts. They included the boundaries and polygons for each zip code and district.

First, we drew a thematic map showcasing the boundaries of the election districts and zip codes. The `combined` map shows the election districts contained in each zip code.

Then, we joined the election district shapefile with the `election` data frame from earlier. We selected the majority winner from each election district. We assigned them a negative value if majority was Trump and a positive value if majority was Biden. This way when we plotted the shapefile, the negative value were red and positive values were blue. The closer it was to 1, the darker it was. Similarly, we joined the zip code shapefile with the `covid` data from earlier to plot the 7 day rolling average positivity rates on the day of election.

Lastly we created the `zip_aded` which was just a data frame with the Zip codes and ADED codes so that we can use it as a reference to join `covid` and `election`.

```{r, warning = FALSE, message = FALSE}
# download.file("https://github.com/data607project3-SMDOJ/finalproject/blob/main/Data/geo_export_1acb0e0a-cfac-4e46-99c6-44540a928d68.shp", "geo_export_1acb0e0a-cfac-4e46-99c6-44540a928d68.shp")
# ed_geo <- st_read("geo_export_1acb0e0a-cfac-4e46-99c6-44540a928d68.shp")

ed_geo <- st_read("C:/Users/Home/Downloads/Election Districts/geo_export_1b77199e-c7e9-4c48-a1db-4b30f4ef79a5.shp")

# zipcode_geo <- st_read("C:/Users/Home/Downloads/ZIP_CODE_040114/ZIP_CODE_040114.shp")

# download.file("https://data.cityofnewyork.us/api/geospatial/pri4-ifjk?method=export&format=Shapefile", "geo_export_1acb0e0a-cfac-4e46-99c6-44540a928d68.shp")
# ny_zc_geo <- st_read("geo_export_1acb0e0a-cfac-4e46-99c6-44540a928d68.shp")

ny_zc_geo <- st_read("C:/Users/Home/Downloads/Modified Zip Code Tabulation Areas (MODZCTA)/geo_export_1acb0e0a-cfac-4e46-99c6-44540a928d68.shp")

qtm(ed_geo) +
  tm_legend(show = FALSE)

qtm(ny_zc_geo) +
  tm_legend(show = FALSE)


tmap_mode("plot")
## tmap mode set to plotting
combined <- tm_shape(ed_geo) +
  tm_borders("red") +
  tm_shape(ny_zc_geo) +
  tm_borders("blue", lwd = .5)  +
  tm_legend(show = FALSE)

combined

ed_filter <- ed_geo %>%
  left_join(election, by = c("elect_dist" = "ADED")) %>%
  group_by(elect_dist) %>%
  filter(Proportion == max(Proportion)) %>%
  mutate(Proportion = ifelse(Candidate == "Trump", Proportion * -1, Proportion))

  
ed_filter %>%
  ggplot(.) +
  geom_sf(aes(fill = Proportion), color = NA) +
  scale_fill_gradient2() +
  ggtitle("Election Results")

#low = "red",mid = scales::muted("purple"), high = "blue"

zip_filter <- left_join(ny_zc_geo, covid, by = c("modzcta" = "zipcode"))

zip_filter %>%
  ggplot(.) +
  geom_sf(aes(fill = rate)) +
  scale_fill_gradient(low = "white", high = "dark red") +
  ggtitle("COVID-19 rates by Zip Code on Election Day")

st_crs(ed_geo) <- 4269 
st_crs(ny_zc_geo) <- 4269

#ed_geo <- st_transform(ed_geo, 4269)
#ny_zc_geo <- st_transform(ny_zc_geo, 4269)

zip_aded <- st_join(ny_zc_geo,  ed_geo, join = st_contains) %>%
  dplyr::select(c(elect_dist, modzcta)) %>%
  st_set_geometry(., NULL) %>%
  set_colnames(c("ADED", "zipcode")) %>%
  na.omit() %>%
  filter(zipcode != "99999")
```

Below is when we used the  `tigris` package which allows us to download TIGER/Line
shapefiles from the US Census Bureau. It does take a few minutes to download the 
data just for New York alone. 

```{r, echo = FALSE}
#The process below takes too long so its reuploaded via csv

#ny_zipcode_geo <- zctas(cb = TRUE, state = "36")

#combined <- st_join(ed_geo,  ny_zipcode_geo, join = st_intersects)

#zipcode_aded <- combined %>%
#  select(c(elect_dist, ZCTA5CE10)) %>%
#  st_set_geometry(., NULL) %>%
#  set_colnames(c("ADED", "zipcode"))

#write.csv(zipcode_aded, "C:/Users/Home/Desktop/MSDS/DATA 607/zipcode_aded.csv")

#zipcode_aded <- read.csv("https://raw.githubusercontent.com/data607project3-SMDOJ/finalproject/main/Data/zipcode_aded.csv") %>%
#  select(c("ADED", "zipcode"))
```


### Data Visualization

Now since we had a data frame to be used as a reference, we joined the `covid` and `election` data. We made a scatterplot to show the relationship between the proportion for each presidential election and the Covid 7-day rolling average positivity rates.

```{r, warning = FALSE, message = FALSE}
covid_election <- left_join(covid, zip_aded, by = "zipcode") %>%
  left_join(., election, by = "ADED") %>%
  na.omit(Propotion)

covid_election <- covid_election[(7:nrow(covid_election)), ]

covid_election %>%
  filter(Candidate == "Biden") %>%
  ggplot(., aes(x = Proportion, y = rate)) + 
  geom_point(color = "Light Blue") + 
  ggtitle("Proportion of Biden Voters vs Covid Rate in NYC by ED") +
  geom_smooth(method = "lm", color = "Black")

covid_election %>%
  filter(Candidate == "Trump") %>%
  ggplot(., aes(x = Proportion, y = rate)) + 
  geom_point(color = "lightcoral") + 
  ggtitle("Proportion of Trump Voters vs Covid Rate in NYC by ED") +
  geom_smooth(method = "lm", color = "Black")
```

```{r, warning = FALSE, message = FALSE}
covid_election %>%
  filter(Candidate == "Biden") %>%
  ggplot(., aes(x = Proportion, y = rate)) + 
  geom_point(color = "Light Blue") + 
  ggtitle("Proportion of Biden Voters vs Covid Rate by County") +
  geom_smooth(method = "lm", color = "Black") +
  facet_wrap(~County)

covid_election %>%
  filter(Candidate == "Trump") %>%
  ggplot(., aes(x = Proportion, y = rate)) + 
  geom_point(color = "lightcoral") + 
  ggtitle("Proportion of Trump Voters vs Covid Rate by County") +
  geom_smooth(method = "lm", color = "Black") +
  facet_wrap(~County)

zip_ce <- covid_election %>%
  dplyr::select(-c(AD, ED,ADED, Proportion)) %>%
  group_by(zipcode, Candidate) %>%
  mutate(Tally = sum(Tally)) %>%
  ungroup() %>%
  unique() %>%
  group_by(zipcode) %>%
  mutate(Proportion = Tally / sum(Tally)) %>%
  filter(sum(Tally) != 0) %>%
  ungroup()

zip_ce %>%
  filter(Candidate == "Biden") %>%
  ggplot(., aes(x = Proportion, y = rate)) + 
  geom_point(color = "Light Blue") + 
  ggtitle("Proportion of Biden Voters vs Covid Rate in NYC by Zipcode") +
  geom_smooth(method = "lm", color = "Black")

zip_ce %>%
  filter(Candidate == "Trump") %>%
  ggplot(., aes(x = Proportion, y = rate)) + 
  geom_point(color = "lightcoral") + 
  ggtitle("Proportion of Trump Voters vs Covid Rate in NYC by Zipcode") +
  geom_smooth(method = "lm", color = "Black")
```

### Multiple Regression Analysis {.tabset}

We also did a multiple regression analysis to see if there is indeed a relationship 
between COVID positivity rates and the election results. we were able to build a 
model with an adjusted $R^2$ of 43.79%, which means that nearly 44% of the COVID 
7 day moving average rates can be explained by the explanatory variables.

```{r, echo = FALSE}
covid_election <- covid_election %>%
  mutate(Candidate = as.factor(Candidate),
         County = as.factor(County))

zip_ce <- zip_ce %>%
  mutate(Candidate = as.factor(Candidate),
         County = as.factor(County))
```

#### Model 1

```{r}
model <- lm(rate ~  County * Candidate + I(Proportion^2), data = covid_election)
summary(model)
```

#### Model 2

```{r}
model2 <- lm(log1p(rate) ~  County + Candidate + log1p(Proportion), data = covid_election)
summary(model2)
```

#### Model 3

```{r}
model3 <- lm(rate ~  County + Candidate * I(Proportion^2), data = covid_election)
summary(model3)
```

#### Model 4

```{r}
model4 <- lm(rate ~  County + Candidate  + Proportion, data = zip_ce)
summary(model4)
```


### Conclusion

In conclusion, we find a connection between higher rates of COVID infection in zip codes where Trump votes were higher than Biden votes in New York City.  This is more evident in Kings (Brooklyn), Richmond (Staten Island), and New York (Manhattan). According to our multiple regression analysis, with 35-40% of the variability in COVID rates explained by the independent variables, we found the variables with Trump as Candidate and Staten Island had an overall positive effect on the COVID-19 positivity rates.

Also there are some limitations to the data as there are other factors that contribute to COVID rates and that some election districts are shared between zip codes. 
